Parsing Postfix Log Files

### 2009-08-27 doc/logparser-acronyms.tex
Turned regex into an acronym everywhere.

\newglossaryentry{regex}{
    type=\acronymtype,
    name={regex},
    description={Regular Expression},
    first={Regular Expression (regex)}
}
\newcommand{\regexglossarydescription}[0]{
    Regular Expressions are a compact, powerful method of specifying
    patterns that describe a set of strings.  The regex \texttt{aa*b*b}
    describes a set of strings, all of which start with \texttt{a},
    followed by any number of \texttt{a}, then any number of \texttt{b},
    and finish with \texttt{b}; the strings \texttt{ab},
    \texttt{aaaaaaaab}, and \texttt{aaabbbbb} are members of that set,
    whereas the strings \texttt{abba}, \texttt{abcd}, and \texttt{qwerty}
    are not.  A string can be checked against a regex to determine if the
    string is a member of the set of strings described by that regex.
}
\newglossaryentry{regex glossary}{
    name={Regular Expression},
    description={
        \regexglossarydescription{}%
    }
}

>>>>

\newglossaryentry{regex}{
    type=\acronymtype,
    name={regex},
    description={Regular Expression},
    first={Regular Expression (regex)}
}
\newcommand{\regexglossarydescription}[0]{
    Regular Expressions are a compact, powerful method of specifying
    patterns that describe a set of strings.  The \acronym{regex}
    \texttt{aa*b*b} describes a set of strings, all of which start with
    \texttt{a}, followed by any number of \texttt{a}, then any number of
    \texttt{b}, and finish with \texttt{b}; the strings \texttt{ab},
    \texttt{aaaaaaaab}, and \texttt{aaabbbbb} are members of that set,
    whereas the strings \texttt{abba}, \texttt{abcd}, and \texttt{qwerty}
    are not.  A string can be checked against a \acronym{regex} to
    determine if the string is a member of the set of strings described by
    that \acronym{regex}.
}
\newglossaryentry{regex glossary}{
    name={Regular Expression},
    description={
        \regexglossarydescription{}%
    }
}
\newcommand{\regexes}[0]{
    \acronym[es]{regex}%
}

### 2009-08-27 doc/logparser-architecture.tex
Changes to capitalisation of titles, as specified by Howard Hamilton. Fixed that/than type identified by Joe Carthy.
Turned regex into an acronym everywhere.

Separating the rules from the actions and framework makes it possible to
parse new inputs without modifying the core parsing algorithm.  Adding a
new rule with the action to invoke and a regex to recognise inputs is
trivial in comparison to understanding an entire parser, identifying the
correct location to change, and making the appropriate changes.  Changes to
a parser must be made without adversely affecting existing parsing,
including any edge cases that are not immediately obvious; an edge case
that occurs only four times in \numberOFlogFILES{} log files is described
in Yet More Aborted Delivery Attempts (\sectionref{yet more aborted
delivery attempts}).  The more intrusive the changes are, the more likely
they are to introduce a bug, so reducing the extent of the changes is
important.  Requiring changes to a parser's source code also complicates
upgrades of the parser, because the changes must be preserved during the
upgrade, and they may clash with changes made by the developer.  This
architecture allows the user to add new rules to a parser without having to
edit it, unless the new inputs cannot be processed by the existing actions.
If the new inputs do require new functionality, new actions can be added to
the parser without having to modify existing actions; only when the new
actions need to cooperate with existing actions will more extensive changes
be required.

@@

            \item Checking that the regex in the rule is valid.

@@

Rules are responsible for recognising inputs: each rule should recognise
one and only one input variant; an input category with multiple input
variants should have multiple rules, one for each variant.  Rules will
typically use a regex when recognising inputs, but other approaches may
prove useful for some applications, e.g.\ comparing fixed strings to the
input, or checking the length of the input; for the remainder of this
thesis it will be assumed that a regex is used.  Each rule must specify the
regex to recognise inputs and the action to invoke when recognition is
successful, but implementations are free to add any other attributes they
require; \sectionref{rules table} describes the attributes used in
\parsername{}, and some generally useful attributes will be discussed later
in this section.

@@

    \item Modify an existing rule's regex, because the new input is part of
        an existing variant.

    \item Write a new rule that pairs an existing action with a new regex,
        adding a new variant to an existing category.

    \item Create a new category of inputs, write a new action to process
        inputs from the new category, and write a new rule pairing the new
        action with a new regex.

@@

\subsection{Attaching Conditions To Rules}

@@

An example of how rule conditions can be used is parsing C-style comments,
which start with \texttt{/*} and end with \texttt{*/}; the start and end
tokens can be on one line, or may have many lines between them.
\Tableref{Rules to parse C-style comments} shows the regexes, conditions,
and state changes of the four rules required to parse C-style comments.
These are simplified rules, e.g.\ rules one and two will incorrectly
recognise the comment start token if it is within a quoted string.  Rules
one and two will be used when the parser is parsing code, not comments:
rule one recognises a comment that is contained within one line and leaves
the parser's state unchanged; rule two recognises the start of a comment
and changes the parser's state to parsing comments instead of parsing code.
Rules three and four will be used when the parser is parsing a comment:
rule three recognises the end of a comment and switches the parser's state
back to parsing code; rule four recognises a comment line without an end
token and keeps the parser's state unchanged.  It is important that the
rules are applied in the order listed in \tableref{Rules to parse C-style
comments} because rule two overlaps with rule one, and rule four overlaps
with rule three; \sectionref{rules in architecture} has explained how this
is achieved.  \sectionref{overlapping rules in architecture} will discuss
the benefits and difficulties of using overlapping rules.

@@

Overlapping rules can be detected by visual inspection, or a program could
be written to analyse the regexes in a ruleset.  Traditional regexes are
equivalent in computational power to \acronym{FA} and can be converted to
\acronym{FA}, so regex overlap can be detected by finding a non-empty
intersection of two \acronym{FA}\@.  Perl 5.10 regexes~\cite{perlre} are
more powerful than traditional regexes: they can match correctly balanced
brackets nested to an arbitrary depth, e.g.\
\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!  matches
\verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires the
regex engine to maintain state on a stack, so Perl 5.10 regexes are
equivalent in computational power to \acronym{PDA}; detecting overlap may
require calculating the intersection of two \acronym{PDA} instead of two
\acronym{FA}.  \acronym{PDA} intersection is not closed, i.e.\ the result
cannot always be implemented using a \acronym{PDA}, so intersection may be
intractable sometimes, e.g.:
$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
Detecting overlap amongst $n$ regexes requires calculating
$\frac{n\left(n-1\right)}{2}$ intersections, resulting in
$O\left(n^{2}x\right)$ complexity, where $x$ is the cost of calculating
\acronym{FA} or \acronym{PDA} intersection.  This is certainly not a task
to be performed every time a parser runs: naive detection of overlap
amongst \parsernames{} \numberOFrules{} rules would require calculating
\numberOFruleINTERSECTIONS{} intersections.

When detecting overlap, any conditions attached to rules must be taken into
account, because two rules whose regexes overlap may have conditions
attached that prevent the rules overlapping.  A less naive approach to
overlap detection would first check for overlapping conditions amongst
rules, and then check for overlap between the regexes of each pair of rules
with overlapping conditions.  Rule overlap is not transitive, e.g.\ given
these three conditions:

@@

    $y$ & = & cost of checking for overlap between two conditions \\
    $o$ & = & set of pairs of rules with overlapping conditions   \\
    $x$ & = & cost of checking for overlap between two regexes    \\

@@

In the worst case, $|o|$ above will be equal to $n^{2}$.  For this approach
to be more efficient than the naive approach, $y$ must be significantly
lower than $x$.  If $y$ is higher than $x$ then the checks for overlap
should be performed in the opposite order: check for regex overlap first,
then check for condition overlap only between pairs of rules with
overlapping regexes.

@@

It is possible to write pathological regexes, which fall into two main
categories: regexes that match inputs they should not, and regexes that
consume excessive amounts of CPU time during matching.  Defining a regex
that matches inputs it should not is trivial: \verb!/^/! matches the start
of every input.  This regex would be found by a tool that detects
overlapping rules, and would easily be noticed by visual inspection, but
more complex regexes would be harder to find.  Regexes that match inputs
more than they should are a problem not because of excessive resource
usage, but because they may prevent the correct rule from recognising the
input.  If an adaptive ordering system is used to prioritise rules that
frequently recognise inputs (see \sectionref{rule ordering for
efficiency}), then a rule with a regex that matches inputs it should not
may be promoted up through the list, displacing an increasing number of
correct rules as it rises.

Excessive CPU time is usually consumed when a regex fails to match an
input, and the regex engine backtracks many times because of alteration or
nested quantifiers; successful matching is generally quite fast with such
regexes, so problematic regexes may go unnoticed for a long time.  For
example, with most regex engines matching double quoted strings with
\verb!\"([^"\\]+|\\.)*"\!  is very fast when a match can be found, but when
the match fails its computational complexity for a string of length $n$ is
$O(2^{n})$; see~\cite{mastering-regular-expressions} for in-depth
discussion of nested quantifiers, backtracking, alteration, and capturing
groups.  Pathological regexes that consume excessive CPU time can be
difficult to detect, whether by visual inspection or by machine inspection,
but if a regex is converted to a \acronym{FA} or the internal
representation used by the regex engine, it may be possible to determine if
nested quantifiers or other troublesome constructs are present.  Modern
regex engines have addressed many of these problems, e.g.\ the regex to
match double quoted strings given above fails immediately with Perl 5.10,
regardless of the input length, because the regex engine looks for both of
the required double quotes first.  Similarly, Perl 5.10's regex engine
optimises alterations starting with literal text into a trie, which has
matching time proportional to the length of the alternatives, rather than
the number of alternatives.  Perl regexes can use \verb!(?>pattern)!, which
matches \verb!pattern! the first time the regex engine passes over it, and
does not change what it originally matched if the regex engine backtracks
over it, alleviating problems caused by excessive backtracking; Prolog
users will notice a similarity to the \verb'!' (cut) operator.  A
presentation showing some of Perl 5.10's new regex features is available at
\urlLastChecked{http://www.regex-engineer.org/slides/perl510_regex.html}{2009/03/03}.

@@

This chapter has presented the parser architecture developed for this
project.  It started with a high level view of the architecture, describing
how it achieves its design aim of being easily extensible for users, and
the advantages that being easily extensible brings to parser authors.  The
three main components of the architecture were documented in detail,
explaining each component's responsibilities and the functionality it
provides, plus any difficulties associated with the components.  The
framework provides several support functions and manages the parsing
process, enabling simple rules and actions to be written.  The actions are
simple to understand, because the architecture does not impose any
structure or requirements upon them: parser authors are free to do anything
they want within an action.  The architecture's support for cascaded
parsing was described in the actions section, with an example to illustrate
how it can be useful for general parsing.  The rules section was the
longest section in this chapter, because although the rules appear to be
quite simple --- recognise an input and specify the action to invoke ---
they have subtle behaviour that needs to be clearly explained.  When
extending a ruleset, a decision needs to be taken about whether the input
should be recognised by extending an existing rule, by adding a new rule to
an existing input category, or by adding a new input category, action, and
rule.  Rules can have conditions attached to them, restricting the set of
rules used to recognise an input; the complexity of the conditions used
greatly influences the difficulty of writing a correct ruleset or
understanding and extending an existing ruleset.  Overlapping rules are
frequently a requirement in a parser, and their use can greatly simplify
some rules, but they can be a source of bugs because they can recognise
inputs unexpectedly.  The framework does not try to detect overlapping
rules, because overlap amongst rules may be valid and is quite often
intentional; that responsibility falls to the author of the ruleset.  The
difficulty involved in detecting overlap is proportional to the complexity
of a ruleset's regexes and conditions, and may be possible for a human yet
intractable or impossible for a program.  The rules section concludes with
a discussion of pathological rules, concentrating on pathological regexes.

>>>>

Separating the rules from the actions and framework makes it possible to
parse new inputs without modifying the core parsing algorithm.  Adding a
new rule with the action to invoke and a \acronym{regex} to recognise
inputs is trivial in comparison to understanding an entire parser,
identifying the correct location to change, and making the appropriate
changes.  Changes to a parser must be made without adversely affecting
existing parsing, including any edge cases that are not immediately
obvious; an edge case that occurs only four times in \numberOFlogFILES{}
log files is described in Yet More Aborted Delivery Attempts
(\sectionref{yet more aborted delivery attempts}).  The more intrusive the
changes are, the more likely they are to introduce a bug, so reducing the
extent of the changes is important.  Requiring changes to a parser's source
code also complicates upgrades of the parser, because the changes must be
preserved during the upgrade, and they may clash with changes made by the
developer.  This architecture allows the user to add new rules to a parser
without having to edit it, unless the new inputs cannot be processed by the
existing actions.  If the new inputs do require new functionality, new
actions can be added to the parser without having to modify existing
actions; only when the new actions need to cooperate with existing actions
will more extensive changes be required.

@@

            \item Checking that the \acronym{regex} in the rule is valid.

@@

Rules are responsible for recognising inputs: each rule should recognise
one and only one input variant; an input category with multiple input
variants should have multiple rules, one for each variant.  Rules will
typically use a \acronym{regex} when recognising inputs, but other
approaches may prove useful for some applications, e.g.\ comparing fixed
strings to the input, or checking the length of the input; for the
remainder of this thesis it will be assumed that a \acronym{regex} is used.
Each rule must specify the \acronym{regex} to recognise inputs and the
action to invoke when recognition is successful, but implementations are
free to add any other attributes they require; \sectionref{rules table}
describes the attributes used in \parsername{}, and some generally useful
attributes will be discussed later in this section.

@@

    \item Modify an existing rule's \acronym{regex}, because the new input
        is part of an existing variant.

    \item Write a new rule that pairs an existing action with a new
        \acronym{regex}, adding a new variant to an existing category.

    \item Create a new category of inputs, write a new action to process
        inputs from the new category, and write a new rule pairing the new
        action with a new \acronym{regex}.

@@

\subsection{Attaching Conditions to Rules}

@@

An example of how rule conditions can be used is parsing C-style comments,
which start with \texttt{/*} and end with \texttt{*/}; the start and end
tokens can be on one line, or may have many lines between them.
\Tableref{Rules to parse C-style comments} shows the \regexes{},
conditions, and state changes of the four rules required to parse C-style
comments.  These are simplified rules, e.g.\ rules one and two will
incorrectly recognise the comment start token if it is within a quoted
string.  Rules one and two will be used when the parser is parsing code,
not comments: rule one recognises a comment that is contained within one
line and leaves the parser's state unchanged; rule two recognises the start
of a comment and changes the parser's state to parsing comments instead of
parsing code.  Rules three and four will be used when the parser is parsing
a comment: rule three recognises the end of a comment and switches the
parser's state back to parsing code; rule four recognises a comment line
without an end token and keeps the parser's state unchanged.  It is
important that the rules are applied in the order listed in \tableref{Rules
to parse C-style comments} because rule two overlaps with rule one, and
rule four overlaps with rule three; \sectionref{rules in architecture} has
explained how this is achieved.  \sectionref{overlapping rules in
architecture} will discuss the benefits and difficulties of using
overlapping rules.

@@

Overlapping rules can be detected by visual inspection, or a program could
be written to analyse the \regexes{} in a ruleset.  Traditional \regexes{}
are equivalent in computational power to \acronym{FA} and can be converted
to \acronym{FA}, so \acronym{regex} overlap can be detected by finding a
non-empty intersection of two \acronym{FA}\@.  Perl 5.10
\regexes{}~\cite{perlre} are more powerful than traditional \regexes{}:
they can match correctly balanced brackets nested to an arbitrary depth,
e.g.\ \verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!  matches
\verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires the
\acronym{regex} engine to maintain state on a stack, so Perl 5.10
\regexes{} are equivalent in computational power to \acronym{PDA};
detecting overlap may require calculating the intersection of two
\acronym{PDA} instead of two \acronym{FA}.  \acronym{PDA} intersection is
not closed, i.e.\ the result cannot always be implemented using a
\acronym{PDA}, so intersection may be intractable sometimes, e.g.:
$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
Detecting overlap amongst $n$ \regexes{} requires calculating
$\frac{n\left(n-1\right)}{2}$ intersections, resulting in
$O\left(n^{2}x\right)$ complexity, where $x$ is the cost of calculating
\acronym{FA} or \acronym{PDA} intersection.  This is certainly not a task
to be performed every time a parser runs: naive detection of overlap
amongst \parsernames{} \numberOFrules{} rules would require calculating
\numberOFruleINTERSECTIONS{} intersections.

When detecting overlap, any conditions attached to rules must be taken into
account, because two rules whose \regexes{} overlap may have conditions
attached that prevent the rules overlapping.  A less naive approach to
overlap detection would first check for overlapping conditions amongst
rules, and then check for overlap between the \regexes{} of each pair of
rules with overlapping conditions.  Rule overlap is not transitive, e.g.\
given these three conditions:

@@

    $y$ & = & cost of checking for overlap between two conditions \\
    $o$ & = & set of pairs of rules with overlapping conditions   \\
    $x$ & = & cost of checking for overlap between two \regexes{}    \\

@@

In the worst case, $|o|$ above will be equal to $n^{2}$.  For this approach
to be more efficient than the naive approach, $y$ must be significantly
lower than $x$.  If $y$ is higher than $x$ then the checks for overlap
should be performed in the opposite order: check for \acronym{regex}
overlap first, then check for condition overlap only between pairs of rules
with overlapping \regexes{}.

@@

It is possible to write pathological \regexes{}, which fall into two main
categories: \regexes{} that match inputs they should not, and \regexes{}
that consume excessive amounts of CPU time during matching.  Defining a
\acronym{regex} that matches inputs it should not is trivial: \verb!/^/!
matches the start of every input.  This \acronym{regex} would be found by a
tool that detects overlapping rules, and would easily be noticed by visual
inspection, but more complex \regexes{} would be harder to find.  Regexes
that match inputs more than they should are a problem not because of
excessive resource usage, but because they may prevent the correct rule
from recognising the input.  If an adaptive ordering system is used to
prioritise rules that frequently recognise inputs (see \sectionref{rule
ordering for efficiency}), then a rule with a \acronym{regex} that matches
inputs it should not may be promoted up through the list, displacing an
increasing number of correct rules as it rises.

Excessive CPU time is usually consumed when a \acronym{regex} fails to
match an input, and the \acronym{regex} engine backtracks many times
because of alteration or nested quantifiers; successful matching is
generally quite fast with such \regexes{}, so problematic \regexes{} may go
unnoticed for a long time.  For example, with most \acronym{regex} engines
matching double quoted strings with \verb!\"([^"\\]+|\\.)*"\!  is very fast
when a match can be found, but when the match fails its computational
complexity for a string of length $n$ is $O(2^{n})$;
see~\cite{mastering-regular-expressions} for in-depth discussion of nested
quantifiers, backtracking, alteration, and capturing groups.  Pathological
\regexes{} that consume excessive CPU time can be difficult to detect,
whether by visual inspection or by machine inspection, but if a
\acronym{regex} is converted to a \acronym{FA} or the internal
representation used by the \acronym{regex} engine, it may be possible to
determine if nested quantifiers or other troublesome constructs are
present.  Modern \acronym{regex} engines have addressed many of these
problems, e.g.\ the \acronym{regex} to match double quoted strings given
above fails immediately with Perl 5.10, regardless of the input length,
because the \acronym{regex} engine looks for both of the required double
quotes first.  Similarly, Perl 5.10's \acronym{regex} engine optimises
alterations starting with literal text into a trie, which has matching time
proportional to the length of the alternatives, rather than the number of
alternatives.  Perl \regexes{} can use \verb!(?>pattern)!, which matches
\verb!pattern! the first time the \acronym{regex} engine passes over it,
and does not change what it originally matched if the \acronym{regex}
engine backtracks over it, alleviating problems caused by excessive
backtracking; Prolog users will notice a similarity to the \verb'!' (cut)
operator.  A presentation showing some of Perl 5.10's new \acronym{regex}
features is available at
\urlLastChecked{http://www.regex-engineer.org/slides/perl510_regex.html}{2009/03/03}.

@@

This chapter has presented the parser architecture developed for this
project.  It started with a high level view of the architecture, describing
how it achieves its design aim of being easily extensible for users, and
the advantages that being easily extensible brings to parser authors.  The
three main components of the architecture were documented in detail,
explaining each component's responsibilities and the functionality it
provides, plus any difficulties associated with the components.  The
framework provides several support functions and manages the parsing
process, enabling simple rules and actions to be written.  The actions are
simple to understand, because the architecture does not impose any
structure or requirements upon them: parser authors are free to do anything
they want within an action.  The architecture's support for cascaded
parsing was described in the actions section, with an example to illustrate
how it can be useful for general parsing.  The rules section was the
longest section in this chapter, because although the rules appear to be
quite simple --- recognise an input and specify the action to invoke ---
they have subtle behaviour that needs to be clearly explained.  When
extending a ruleset, a decision needs to be taken about whether the input
should be recognised by extending an existing rule, by adding a new rule to
an existing input category, or by adding a new input category, action, and
rule.  Rules can have conditions attached to them, restricting the set of
rules used to recognise an input; the complexity of the conditions used
greatly influences the difficulty of writing a correct ruleset or
understanding and extending an existing ruleset.  Overlapping rules are
frequently a requirement in a parser, and their use can greatly simplify
some rules, but they can be a source of bugs because they can recognise
inputs unexpectedly.  The framework does not try to detect overlapping
rules, because overlap amongst rules may be valid and is quite often
intentional; that responsibility falls to the author of the ruleset.  The
difficulty involved in detecting overlap is proportional to the complexity
of a ruleset's \regexes{} and conditions, and may be possible for a human
yet intractable or impossible for a program.  The rules section concludes
with a discussion of pathological rules, concentrating on pathological
\regexes{}.

### 2009-08-27 doc/logparser-conclusion.tex
Turned regex into an acronym everywhere.

Parsing Postfix log files appears at first sight to be an uncomplicated
task, especially if one has previous experience in parsing log files, but
it turns out to be a much more taxing project than initially expected.  The
variety and breadth of log lines produced by Postfix is quite surprising,
because a quick survey of sample log files gives the impression that the
number of distinct log line variants is quite small; this mistaken
impression comes from the uneven distribution exhibited by log lines
produced in normal operation, vividly illustrated in \graphref{rule hits
graph}.  Given the diverse nature of Postfix log lines, and the ease with
which administrators can cause new log lines to be logged
(\sectionref{postfix background}), enabling users to easily extend the
parser to deal with new log lines is a design imperative
(\sectionref{parser design}).  Providing a tool to ease the generation of
regexes from unrecognised log lines (\sectionref{creating new rules in
implementation}) should greatly help users who need to extend their ruleset
to recognise previously unrecognised log lines.

>>>>

Parsing Postfix log files appears at first sight to be an uncomplicated
task, especially if one has previous experience in parsing log files, but
it turns out to be a much more taxing project than initially expected.  The
variety and breadth of log lines produced by Postfix is quite surprising,
because a quick survey of sample log files gives the impression that the
number of distinct log line variants is quite small; this mistaken
impression comes from the uneven distribution exhibited by log lines
produced in normal operation, vividly illustrated in \graphref{rule hits
graph}.  Given the diverse nature of Postfix log lines, and the ease with
which administrators can cause new log lines to be logged
(\sectionref{postfix background}), enabling users to easily extend the
parser to deal with new log lines is a design imperative
(\sectionref{parser design}).  Providing a tool to ease the generation of
\regexes{} from unrecognised log lines (\sectionref{creating new rules in
implementation}) should greatly help users who need to extend their ruleset
to recognise previously unrecognised log lines.
